Balsa: A Compact Line Notation Based on SMILES

### 2022-07-31 manuscript.tex
Revises section headings. Closes #35
Unifies use of the terms node and edge. Closes #36

\section*{Goals and Overall Design}

@@

@@

\section*{Molecular Tree}
\label{molecular-tree}

@@

@@

Gaps and bridges decouple connectivity from bonding, but for opposite purposes. Whereas a gap connects two atoms without bonding them, a bridge creates a bond between two atoms without connecting them. These relationships can complicate the analysis of connectivity and bonding. For example, the atom succeeding a gap behaves in most situations as if it were a root node. When tallying the bonds surrounding such a node, only children are considered. Similar considerations apply when a gap succeeds an atom. Likewise, a bridge node behaves as if it were an atom in the sense that it connects to an atom through a bond. Nevertheless, a bridge node merely stands in for another atom found elsewhere in the tree. Bonding and connectivity analyses within molecular trees must therefore account for the possibility of connection-free bonds and bond-free connections.

\section*{Atoms and Bonds}
\label{atoms-and-bonds}

@@

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation. Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting. However, not all atoms are eligible to do so (see: \nameref{computing-implicit-hydrogen-count}).

@@

\section*{Electron Counting}
\label{electron-counting}

@@

\section*{Delocalization Subgraph}
\label{delocalization-subgraph}

@@

The opposite operation can be accomplished with a \textit{selection algorithm}. A selection algorithm selects two or more nodes, thereby adding them to the DS. The only requirement for a selection algorithm is that the resulting DS must have a perfect matching. Depending on the application, other criteria may be applied. For example, a selection algorithm may restrict candidate atoms to those found in cycles. Electron-counting techniques may also be introduced to approximate the chemical concept of \enquote{aromaticity.} Elision of the single bonds to be added to the DS completes the process.

\section*{Valence and Subvalence}
\label{valence-and-subvalence}

@@

\section*{Computing Implicit Hydrogen Count}
\label{computing-implicit-hydrogen-count}

@@

\section*{Conformation}
\label{conformation}

@@

A PPB is designated by setting its \ttt{direction} attribute to a non-default variant (\ttt{Up} or \ttt{Down}). The names of these variants refer to a two-dimensional geometrical model in which a parent node and its child are placed along a vertical axis. If the child lies above the parent, the \ttt{Up} variant is used. If the child lies below the parent, the \ttt{Down} variant is used.

@@

\section*{Configuration}
\label{configuration}

@@

@@

\subsection*{Atom}

@@

\section*{Reading Strings}
\label{reading-strings}

@@

\section*{Pruning}

@@

\section*{Writing Strings}

@@

Writers must ensure that all selected atoms can be deselected. Consider pyrrole, erroneously encoded with a selected nitrogen atom (e.g., \enquote{n1cccc1}). The nitrogen atom can not be pruned because of its non-zero subvalence (3 - 2). The DS therefore contains all five atoms and all five bonds. A perfect matching does not exist. A reader receiving such a string must report an error. To avoid this outcome, a writer can consume the subvalence by adding a virtual hydrogen (e.g., \enquote{[nH]1cccc1}). The subvalence of nitrogen in this case is zero, so the atom can be pruned. Doing so leaves a DS with four nodes, four edges, and a perfect matching. Readers will therefore consider the string valid. The larger question of gratuitous selection nevertheless remains.

\section*{Working with Molecular Graphs}
\label{working-with-molecular-graphs}

>>>>

\section*{Goals}

@@

\section*{Molecular Representation}
\label{molecular-representation}

@@

@@

\subsection*{Molecular Tree}
\label{molecular-tree}

@@

Gaps and bridges decouple connectivity from bonding, but for opposite purposes. Whereas a gap connects two atoms without bonding them, a bridge creates a bond between two atoms without connecting them. These relationships can complicate the analysis of connectivity and bonding. For example, the atom succeeding a gap behaves in most situations as if it were a root. When tallying the bonds surrounding such an atom, only children are considered. Similar considerations apply when a gap succeeds an atom. Likewise, a bridge node behaves as if it were an atom in the sense that it connects to an atom through a bond. Nevertheless, a bridge node merely stands in for another atom found elsewhere in the tree. Bonding and connectivity analyses within molecular trees must therefore account for the possibility of connection-free bonds and bond-free connections.

\subsection*{Atoms and Bonds}
\label{atoms-and-bonds}

@@

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen atom and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation. Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting. However, not all atoms are eligible to do so (see: \nameref{computing-implicit-hydrogen-count}).

@@

\subsection*{Electron Counting}
\label{electron-counting}

@@

\subsection*{Delocalization Subgraph}
\label{delocalization-subgraph}

@@

The opposite operation can be accomplished with a \textit{selection algorithm}. A selection algorithm selects two or more atoms, thereby adding them to the DS. The only requirement for a selection algorithm is that the resulting DS must have a perfect matching. Depending on the application, other criteria may be applied. For example, a selection algorithm may restrict candidate atoms to those found in cycles. Electron-counting techniques may also be introduced to approximate the chemical concept of \enquote{aromaticity.} Elision of the single bonds to be added to the DS completes the process.

\subsection*{Valence and Subvalence}
\label{valence-and-subvalence}

@@

\subsection*{Computing Implicit Hydrogen Count}
\label{computing-implicit-hydrogen-count}

@@

\subsection*{Conformation}
\label{conformation}

@@

A PPB is designated by setting its \ttt{direction} attribute to a non-default variant (\ttt{Up} or \ttt{Down}). The names of these variants refer to a two-dimensional geometrical model in which a parent atom and its child are placed along a vertical axis. If the child lies above the parent, the \ttt{Up} variant is used. If the child lies below the parent, the \ttt{Down} variant is used.

@@

\subsection*{Configuration}
\label{configuration}

@@

\subsection*{Grammar}

@@

\subsubsection*{Atom}

@@

\section*{Implementation}

\subsection*{Reading Strings}
\label{reading-strings}

@@

\subsection*{Pruning}

@@

\subsection*{Writing Strings}

@@

Writers must ensure that all selected atoms can be deselected. Consider pyrrole, erroneously encoded with a selected nitrogen atom (e.g., \enquote{n1cccc1}). The nitrogen atom can not be pruned because of its non-zero subvalence (3 - 2). The DS therefore contains all five atoms and all five bonds. A perfect matching does not exist. A reader receiving such a string must report an error. To avoid this outcome, a writer can consume the subvalence by adding a virtual hydrogen (e.g., \enquote{[nH]1cccc1}). The subvalence of nitrogen in this case is zero, so the atom can be pruned. Doing so leaves a DS with four atoms, four edges, and a perfect matching. Readers will therefore consider the string valid. The larger question of gratuitous selection nevertheless remains.

\subsection*{Working with Molecular Graphs}
\label{working-with-molecular-graphs}
